---
title: "Identifying fishing trips and activities"
author: "Tania Mendo"
date: '2022-06-27'
output: html_document
editor_options: 
  chunk_output_type: console
---

#Libraries

```{r libraries, include=FALSE,eval = FALSE}


library(tidyverse)
library(leaflet)
library(leaflet.extras)
library(viridis)
library(sf)

```

#Parameters
Here we specify the threshold values for each parameter. These parameters will change depending on the fishery and require careful thinking by researchers.

```{r parameters, include=FALSE,eval = FALSE}

#Parameters

#to define study area
lat_max <- 59 #degrees
lat_min <- 56 #degrees
lon_max <- -6 #degrees
lon_min <- -8 #degrees

crs_original <- 4326
crs_utm <- 32630
land_buffer <- -10 #metres
speed_filter <- 25#knots
port_buffer <- 200 #metres

```

#1. Define study area

Tracking data might contain positions outside the study area due to many reasons. We have to first remove these data points. You will need to specify roughly your study area.

```{r cleaning, include=FALSE,eval = FALSE}

#**************************
#1.- Define study area
#**************************
df <- read.csv("test_harbours.csv")#59413

df <- df %>%
  filter(latitude>lat_min&latitude <lat_max)

df <- df %>%
  filter(longitude<lon_max&longitude>lon_min)#56642

```

#2. Remove duplicates

Sometimes duplicates are generated while tracking.

```{r duplicates, include=FALSE,eval = FALSE}

df <- df %>% 
  distinct(longitude,latitude, time_stamp,.keep_all = TRUE)

df <- df %>% 
  distinct(longitude,latitude, device_id,.keep_all = TRUE)#56267

#visualise (optional)

wardpal <- colorFactor(viridis(8), df$device_id)

 leaflet() %>%
   addTiles() %>%
     addCircleMarkers(lng = ~longitude,
              lat = ~latitude,
              radius=1,
              popup = ~ as.character(time_stamp),
              label = ~as.character(time_stamp),
              fillColor=~wardpal(device_id),
              fillOpacity = 1,
              stroke=FALSE,
              data = df) %>%
    addScaleBar()%>%
   addLegend(position="bottomright",pal=wardpal,values=df$device_id)

```

#3. Remove points on land

Please be aware you will need a high resolution coastline shapefile to adequately remove points on land.

#AIS
#GNSS - 
#human error

Map from: Contains OS data © Crown copyright and database right 2022. Retrieved from: https://osdatahub.os.uk/downloads/open


```{r points on land, include=FALSE,eval = FALSE}

df_sf = st_as_sf(df, coords = c("longitude", "latitude"), crs = crs_original)
df_sf_utm <- st_transform(df_sf, crs_utm)
coords_utm <- as.data.frame(st_coordinates(df_sf_utm))
df$x_utm <- coords_utm$X
df$y_utm <- coords_utm$Y

coastline <- read_sf(dsn = ".", layer = "OutHebrides")#in UTM
coastline_utm <- st_transform(coastline, crs_utm)
coastline_utm_land <- coastline_utm%>%
  st_buffer(dist=land_buffer)

#try parallel running?
land <-lengths(st_intersects(df_sf_utm, 
                             coastline_utm_land, 
                             parallel = 5)) > 0#
df2<-df[!c(land),]#remove points on land #56237

```


#4. Remove unrealistic velocities

We have to remove records associated with very high speed that are unrealistic.

```{r speed, include=FALSE,eval = FALSE}

options(digits.secs = 3)
df2$time_stamp<-as.POSIXct((df2$time_stamp), format = "%Y-%m-%d %H:%M:%OS",origin="1970-01-01", tz = 'Europe/London')#
df2<- df2[order(df2$device_id, df2$time_stamp),]

#add distances in x and y
df2<-df2%>%
  group_by(device_id)%>%
  mutate(dx=c(0,abs(diff(x_utm))),dy=c(0,abs(diff(y_utm))),dt=c(0,abs(diff(time_stamp))))

df2<-df2%>%
  group_by(device_id)%>%
  mutate(dist=sqrt(dx^2+dy^2))

df2$speed<-df2$dist/df2$dt*1.943
df2<- df2%>% rowid_to_column("seq") 

#loop to remove erroneous speeds

repeat { 
  subset<-df2%>%
    filter(speed>speed_filter)

sel<-factor(subset$seq)

nrows<-length(sel)

  if (nrows==0) {
    break
    } 
  else
df2<-df2[!df2$seq %in% sel,] 

df2<-df2%>%
  group_by(device_id)%>%
  mutate(dx=c(0,abs(diff(x))),dy=c(0,abs(diff(y))),dt=c(0,abs(diff(time_stamp))))

df2<-df2%>%
  group_by(device_id)%>%
  mutate(dist=sqrt(dx^2+dy^2),dt=c(0,abs(diff(time_stamp))))
df2$speed<-df2$dist/df2$dt*1.943

} #56051 obs

```


#5. Define trips

Cut track data into potential trips. Again for this step, knowledge on the fishery is required, and we suggest three potential approaches to define a trip:

1) Based on daily patterns of fishing. In some fisheries, trips are conducted only during daytime. In this case it might be worth "cutting" the track data at midnight to separate the different trips. 

2) Based on leaving and entering specific harbours/ports/jetties. If you have a list of these, you can create a spatial buffer (metres) around them and identify when the vessel is leaving and entering this buffer each time. Points located outside a buffer in between points inside a buffer define a specific trip.If you dont have a comprehevensive list, then cluster analysis to define possible harbours. Sometimes we might not have all information on harbours, or jetties used by small-scale fishers. In this case, XXX Anna

3) Logbook data: Jonathan?

```{r define trips, include=FALSE,eval = FALSE}

#1)Based on fishing activity - daily patterns

df2$date<-as.Date(format(as.POSIXct(df2$time_stamp), "%Y-%m-%d"))
df2$trip_id<-paste(df2$device_id,df2$date)

#2) With harbour list - either dinfe harbours with clusters or you already have a list of harbours

 
#  
```

#6. Delete observations "in port"

Depending on the device you are using, a lot of geospatial data might be collected when a vessel is in port (or an equivalent place such as a mooring, launching beach, etc). To reduce the amount of data collected, it might make sense to delete these observations.  

There are two approaches to identify these points. 

If you have only a few trips, and you do not know where the landing sites are, you might want to delete observations at the beginning and at the end of a trip. For example, you can set a specific distance from the coastline from which you expect a vessel to have left their landing sites and steaming to fishing grounds already.


```{r beg and end, include=FALSE,eval = FALSE}

library(spatstat.utils)

df3<-df2
df3<- df3[order(df3$trip_id, df3$time_stamp),]

df3<-df3%>%
  group_by(trip_id)%>%
  mutate(cum_dist=cumsum(dist), rev_cum_dist=revcumsum(dist))

subset<-df3%>%
  filter(cum_dist<400|rev_cum_dist<400)

coastline_buffer<-coastline_utm%>%
  st_buffer(dist=200)

df_sf = st_as_sf(subset, coords = c("x_utm", "y_utm"), crs = crs_utm)

points_in_coast_buffer <-lengths(st_intersects(df_sf, coastline_buffer)) > 0

df3_sf = st_as_sf(df3, coords = c("x_utm", "y_utm"), crs = crs_utm)
df4<-df_sf[c(points_in_coast_buffer),]#9075

#remove sequence of points in df4

sel<-df4$seq

df2<-df2[!df2$seq %in% sel,] #55256

```

#7.- Remove "spurious" trips

at this point, you might want to check summary statistics for your trips and define again, a sensible criteria to define what are likely trips.

```{r trips, include=FALSE}

trips_summary<-df2%>%
  group_by(trip_id) %>%
  summarise(dist_travelled = sum(dist)/1000,time_travelled=sum(dt/3600,na.rm=TRUE), nobs=n_distinct(time_stamp))  

ggplot()+geom_histogram(data=trips_summary,aes(x=dist_traveled))+theme_bw()#in meters

ggplot()+geom_histogram(data=trips_summary,aes(x=time_travelled))+theme_bw()#in hours

trips_summary<-trips_summary%>%
  filter(time_travelled<1|nobs<50|dist_travelled<5)

sel<-trips_summary$trip_id

df2<-df2[!df2$trip_id %in% sel,] #54986

write.table(df2,"clean_track_data.txt")


```




