---
title: "Identifying fishing trips and activities"
author: "Tania Mendo"
date: '2022-06-27'
output: html_document
editor_options: 
  chunk_output_type: console
---

#Libraries

```{r libraries, include=FALSE,eval = FALSE}

library(tidyverse)
library(sf)
library(spatstat.utils)

```

#Dataset

The dataset should include 4 columns: latitude (WGS84), longitude (WGS84), time stamp (Y-m-d H:M:S) and a device or vessel identifier.

#Parameters
Here we specify the threshold values for each parameter. These parameters will change depending on where and how the fishery operates. Expert knowledge is required to assign a speed threshold and the buffer that will be assigned around each port.

```{r parameters, include=FALSE,eval = FALSE}

#Parameters

#to define study area
lat_max<-59 #degrees
lat_min<-56 #degrees
lon_max<--6 #degrees
lon_min<--8 #degrees

crs_original<-4326 # modify as needed
crs_utm<-32630 # modify as needed
#land_buffer<--10 #- 10 metres
speed_filter<-25#greatest speed possible for vessels during steaming (in knots)
port_buffer<-200 #buffer radius around the port, where points will be removed before inferring fishing activities (in metres)

```

#1. Define study area

Tracking data might contain positions outside the study area due to many reasons. We have to first remove these data points. You will need to specify roughly your study area.

```{r study area, include=FALSE,eval = FALSE}

#**************************
#1.- Define study area
#**************************
df<-read.csv("track_data_202107.csv")
df<-head(df,10000)
df<-df%>%
  filter(latitude>lat_min&latitude <lat_max)

df<-df%>%
  filter(longitude<lon_max&longitude>lon_min)

```

#2. Remove duplicates

```{r duplicates, include=FALSE,eval = FALSE}

df<-df%>% 
  distinct(longitude,latitude, time_stamp,.keep_all = TRUE)

```

#3. Remove points on land

Please be aware you will need a high resolution coastline shapefile to adequately remove points on land.

The current map was downlaoded from: OS data © Crown copyright and database right 2022. Retrieved from: https://osdatahub.os.uk/downloads/open


```{r points on land, include=FALSE,eval = FALSE}

df_sf = st_as_sf(df, coords = c("longitude", "latitude"), crs = crs_original)
df_sf_utm<-st_transform(df_sf, crs_utm)#transform to UTM system
coords_utm<-as.data.frame(st_coordinates(df_sf_utm))#extract coordinates
df$x_utm<-coords_utm$X
df$y_utm<-coords_utm$Y

coastline <- read_sf(dsn = ".", layer = "OutHebrides")#read map
coastline_utm<-st_transform(coastline, crs_utm)

land <-lengths(st_intersects(df_sf_utm, coastline_utm,parallel = 5)) > 0#
df<-df[!c(land),]#remove points on land 

```

#4. Remove unrealistic velocities

We have to remove records associated with very high vessel speeds that are unrealistic.

```{r speed, include=FALSE,eval = FALSE}

options(digits.secs = 3)
df$time_stamp<-as.POSIXct((df$time_stamp), format = "%Y-%m-%d %H:%M:%OS",origin="1970-01-01", tz = 'Europe/London')#
df<- df[order(df$device_id, df$time_stamp),]

#add distances between consecutive observations in x and y
df<-df%>%
  group_by(device_id)%>%
  mutate(dx=c(0,abs(diff(x_utm))),dy=c(0,abs(diff(y_utm))),dt=difftime(strptime(time_stamp, "%Y-%m-%d %H:%M:%S"), lag(strptime(time_stamp, "%Y-%m-%d %H:%M:%S"), units = 'secs')),
         dt = as.numeric(dt, units = 'secs'))%>%
  mutate(dist=sqrt(dx^2+dy^2))%>%
  mutate(speed=dist/dt*1.943)#speed in knots

df<- df%>% rowid_to_column("seq") #add a unique identifier to each row

#loop to remove erroneous speeds

repeat { 
  subset<-df%>%
    filter(speed>speed_filter)

sel<-factor(subset$seq)

nrows<-length(sel)

  if (nrows==0) {
    break
    } 
  else
df<-df[!df$seq %in% sel,] 

df<-df%>%
  group_by(device_id)%>%
  mutate(dx=c(0,abs(diff(x_utm))),dy=c(0,abs(diff(y_utm))),dt=c(0,abs(diff(time_stamp))))

df<-df%>%
  group_by(device_id)%>%
  mutate(dist=sqrt(dx^2+dy^2),dt=c(0,abs(diff(time_stamp))))
df$speed<-df$dist/df$dt*1.943

}

```

#5. Define individual trips

```{r define trips, include=FALSE,eval = FALSE}

#1)Based on fishing activity - daily patterns

df$date<-as.Date(format(as.POSIXct(df$time_stamp), "%Y-%m-%d"))
df$trip_id<-paste(df$device_id,df$date)

# 2) Based on mooring/anchorage sites

#See Sup. Mat.XX R Code "Identifying mooring sites"

#3) From logbook data 
log_book <- read.csv("log_data.csv")

setDT(df2)
setDT(log_book)

#unique trip start and end, with times... (maybe cases for NA needs handling ?)
trips <- log_book[ ,. (id = unique(log_book_nr)),
               by = .(device_id, trip_start, trip_end)]

#use only vessels where we have gps data 
trips <- trips[trips$device_id %in% df2$device_id, ]

#some formatting for "foverlaps", no duplicate column names allowed, but start and end both set to postime
df2$start <- df2$time_stamp
df2$end <- df2$time_stamp

setkey(df2, device_id, start, end)
setkey(trips, device_id, trip_start, trip_end)

## use the foverlaps function to determine where the times overlaps/ where there is sailing
df2 <- foverlaps(df2, trips)

# sailing occurs when trip start is not NA/ when an overlap was found
df2 <- df2[! is.na(df2$trip_start), ]

#make unique trip id
df2$trip_id <- paste0(df2$device_id, df2$id)

```

#6. Delete observations "in port"

```{r ports, include=FALSE,eval = FALSE}

df<- df[order(df$trip_id, df$time_stamp),]

df<-df%>%
  group_by(trip_id)%>%
  mutate(cum_dist=cumsum(dist), rev_cum_dist=revcumsum(dist))

subset<-df%>%
  filter(cum_dist<800|rev_cum_dist<800)

coastline_buffer<-coastline_utm%>%
  st_buffer(dist=port_buffer)

df_sf = st_as_sf(subset, coords = c("x_utm", "y_utm"), crs = crs_utm)

points_in_coast_buffer <-lengths(st_intersects(df_sf, coastline_buffer)) > 0

df_sf<-df_sf[c(points_in_coast_buffer),]#9075

#remove sequence of points in df_sf

sel<-df_sf$seq

df<-df[!df$seq %in% sel,] #55256

```

#7.- Remove "spurious" trips


```{r trips, include=FALSE}

trips_summary<-df%>%
  group_by(trip_id) %>%
  summarise(dist_travelled = sum(dist)/1000,time_travelled=sum(dt/3600,na.rm=TRUE), nobs=n_distinct(time_stamp))  

ggplot()+geom_histogram(data=trips_summary,aes(x=dist_travelled))+theme_bw()#in meters

ggplot()+geom_histogram(data=trips_summary,aes(x=time_travelled))+theme_bw()#in hours

trips_summary<-trips_summary%>%
  filter(time_travelled<1|nobs<50|dist_travelled<5)

sel<-trips_summary$trip_id

df<-df[!df$trip_id %in% sel,] #54986

write.table(df,"clean_trips_data_subset.txt",sep=",")


```

#Inferring fishing activities

1) Data pre-processing

```{r cut trips, include=FALSE}

df2<-read.table("clean_trips_data_subset.txt",sep=",")

options(digits.secs = 3)
df2$time_stamp<-as.POSIXct((df2$time_stamp), format = "%Y-%m-%d %H:%M:%OS",origin="1970-01-01", tz = 'Europe/London')#
df2<- df2[order(df2$trip_id, df2$time_stamp),]

#add distance and time between consecutive observations in x and y
df2<-df2%>%
  group_by(trip_id)%>%
  mutate(dx=c(0,abs(diff(x_utm))),dy=c(0,abs(diff(y_utm))),dt = difftime(strptime(time_stamp, "%Y-%m-%d %H:%M:%S"), lag(strptime(time_stamp, "%Y-%m-%d %H:%M:%S"), units = 'secs')),
         dt = as.numeric(dt, units = 'secs'))%>%
   mutate(dist=sqrt(dx^2+dy^2))

#df2$check<-ifelse(df2$dist>500,1,2)

df2<-df2 %>% 
 group_by(trip_id) %>%
 mutate(newID = ifelse(dist>500, 1,0))#if the distance between two consecutive observations is greater than 500 metres, assign a 1, otherwise a 0.

df2$newID<-cumsum(c(na.omit(df2$newID,na.rm=TRUE)))#add unique identifier for each segment of trip

```

#Regularise tracks

```{r interpolation, include=FALSE}

#create data frame with time_stamps every minute to interpolate positions

interp_df<-df2 %>%
  group_by(newID,trip_id)%>%
  expand(time_stamp = seq(
    from = min(time_stamp),
    to = max(time_stamp),
    by = 'min')) %>%
  arrange(time_stamp) 
#

t <- df2$time_stamp #starting time_stamps
y <- df2$x_utm #longitudes
f <- approxfun(t,y)

interp_df$x_utm<-f(interp_df$time_stamp)
y <- df2$y_utm
f <- approxfun(t,y)
interp_df$y_utm<-f(interp_df$time_stamp)


```

#Remove points on land


```{r points on land, include=FALSE,eval = FALSE}

df_sf = st_as_sf(interp_df, coords = c("x_utm", "y_utm"), crs = crs_utm)

land <-lengths(st_intersects(df_sf, coastline_utm,parallel = 5)) > 0#
interp_df<-interp_df[!c(land),]#remove points on land 

```


#Infer fishing activities

```{r fishing, include=FALSE}

#When no validation data available:
#See Code provided in Mendo et al, 2019 for speed thresholds, Hidden Markov models and Expectation Maximisation algorithms. - 

#Random forests?

#???ANY OTHER???

```

#Remove spurious fishing activities

```{r fishing check, include=FALSE}

interp_df<- interp_df[order(interp_df$newID, interp_df$time_stamp),]

interp_df$rf_behaviour_1<-dplyr::lead(interp_df$rf_behaviour,n=1)
interp_df$rf_behaviour_2<-dplyr::lead(interp_df$rf_behaviour,n=2L)
interp_df$rf_behaviour__1<-dplyr::lag(interp_df$rf_behaviour,n=1L)
interp_df$rf_behaviour__2<-dplyr::lag(interp_df$rf_behaviour,n=2L)

interp_df$fishing<-ifelse(interp_df$rf_behaviour=="not_fishing"&interp_df$rf_behaviour_1=="fishing"&interp_df$rf_behaviour_2=="fishing"&interp_df$rf_behaviour__1=="fishing"&interp_df$rf_behaviour__2=="fishing", "fishing", as.character(interp_df$rf_behaviour))

interp_df$fishing<-ifelse(interp_df$rf_behaviour=="fishing"&interp_df$rf_behaviour_1=="not_fishing"&interp_df$rf_behaviour_2=="not_fishing"&interp_df$rf_behaviour__1=="not_fishing"&interp_df$rf_behaviour__2=="not_fishing", "not_fishing",as.character( interp_df$fishing))


```

#Identify fishing trips

```{r fishing trips, include=FALSE}

fishing_trips<-interp_df%>%
  filter(fishing=="fishing")%>%
  summarise(fishing_trip_ID=unique(trip_id))

```


