---
title: "Identifying mooring/anchoring sites"
author: "Tania Mendo"
date: "2022-08-29"
output: html_document
editor_options: 
  chunk_output_type: console
---

#Identifying harbours or mooring/anchorage sites

```{r libraries, include=FALSE,eval = FALSE}

library(tidyverse)
library(sf)
library(leaflet)#for visualisation
library(leaflet.extras)#for visualisation
library(viridis)#for visualisation
library(cluster)
library(car)
library(data.table)

#Specifications
crs_utm<-32630
crs_original<-4326
buffer_land<-200
distance_moved<-200
time_elapsed<-2*60*60
centroids_buff<-200
```

#read data

Use raw data that is representative of each vessel's activity throughout a year if available (as operational patterns might change throughout the year). 

Input data needed:

vessel identifier
longitude (UTM)
latitude (UTM)
time stamp

```{r data prep, include=FALSE,eval = FALSE}

df<-read.csv("ArtHarbList_Test_5more.csv")# 
df$time_stamp<-as.POSIXct((df$time_stamp), format = "%Y-%m-%d %H:%M:%OS",origin="1970-01-01", tz = 'Europe/London')#
df<- df[order(df$vessel_pln, df$time_stamp),]

#add distances in x and y
df<-df%>%
  group_by(vessel_pln)%>%
  mutate(dx=c(0,abs(diff(x_UTM))),dy=c(0,abs(diff(y_UTM))))%>%
 mutate(dt = difftime(strptime(time_stamp, "%Y-%m-%d %H:%M:%S"), lag(strptime(time_stamp, "%Y-%m-%d %H:%M:%S"), units = 'secs')),
         dt = as.numeric(dt, units = 'secs'))%>%
  mutate(dist=sqrt(dx^2+dy^2))%>%#distance between consecutive observations
  mutate(speed=dist/dt*1.943)#speed between consecutive observations (in knots)

df<- df%>% rowid_to_column("seq") #add a unique identifier

unique(df$vessel_pln)

df<-df%>%
  filter(vessel_pln=="CY146")#select one vessel - check which one is going to be OK for example

```

#Select low velocity points

```{r speed, include=FALSE,eval = FALSE}

# only select observations with low speeds (these locations would be presumably in port, when vessel is not moving) 
 df_1<-df%>%
  filter(speed<0.1)#Subset to speeds less than 0.1 knot
```


#Subset data to observations close to the coastline

```{r coastline, include=FALSE,eval = FALSE}

coastline <- read_sf(dsn = ".", layer = "OutHebrides")#coastline shapefile
coastline_utm<-st_transform(coastline, crs_utm)

df_sf = st_as_sf(df_1, coords = c("x_UTM", "y_UTM"), crs = crs_utm)

#Remove points further away from land

coastline_utm_land<-coastline_utm%>%
  st_buffer(dist = buffer_land) # choose a buffer distance that takes into account how close to land you consider the vessel might be mooring.

land <-lengths(st_intersects(df_sf, coastline_utm_land)) > 0#
df_1<-df_1[c(land),]#remove points outside the buffer

#at this point it is recommended to visualise the remaining positions - in leaflet lat/lons (WGS84) are required

df_sf = st_as_sf(df_1, coords = c("x_UTM", "y_UTM"), crs = crs_utm)
df_sf<-st_transform(df_sf, crs_original)#
coords<-as.data.frame(st_coordinates(df_sf))#extract coordinates
df_1$lon<-coords$X
df_1$lat<-coords$Y
# 
df_1$date<-as.Date(df_1$time_stamp)# do i need this?
df_1$trip_id<-paste(df_1$vessel_pln,df_1$date)# do i need this?

wardpal <- colorFactor(viridis(20), df_1$trip_id)
#
  leaflet() %>%
     #addProviderTiles(providers$CartoDB.Positron)%>% 
     addTiles(urlTemplate = "https://mts1.google.com/vt/lyrs=s&hl=en&src=app&x={x}&y={y}&z={z}&s=G", attribution = 'Google') %>%
      addCircleMarkers(lng = ~lon,
               lat = ~lat,
               radius=5,
               popup = ~ as.character(date),
               label = ~as.character(trip_id),
               fillColor=~wardpal(trip_id),
               fillOpacity = 3,
               stroke=FALSE,
               data = df_1) %>%
     addScaleBar()%>%
    addLegend(position="bottomright",pal=wardpal,values=df_1$trip_id)
```


#Interpolate observations

```{r interpolation, include=FALSE,eval = FALSE}

df_1<- df_1[order(df_1$vessel_pln, df_1$time_stamp),]

df_1<-df_1%>%
  group_by(vessel_pln)%>%
  mutate(dx=c(0,abs(diff(x_UTM))),dy=c(0,abs(diff(y_UTM))))%>%
   mutate(dist=sqrt(dx^2+dy^2))%>%
 mutate(dt = difftime(strptime(time_stamp, "%Y-%m-%d %H:%M:%S"), lag(strptime(time_stamp, "%Y-%m-%d %H:%M:%S"), units = 'secs')),
         dt = as.numeric(dt, units = 'secs'))

df_1<-df_1 %>% 
 group_by(vessel_pln) %>%
 mutate(newID = ifelse(dist>distance_moved|dt>time_elapsed, 1,0 ))#cut into segments if the distance between consecutive observations is greater than 200 metres or if the time between consecutive observations is greater than 2 hours.

df_1$newID2<-ifelse(df_1$newID==1,cumsum(c(0,na.omit(df_1$newID,na.rm=TRUE))),df_1$newID)

df_1$newID3<-cumsum(c(0,na.omit(df_1$newID2,na.rm=TRUE)))

interp_df<-df_1 %>%
  group_by(newID3)%>%
  expand(time_stamp = seq(
    from = min(time_stamp),
    to = max(time_stamp),
    by = 'min')) %>%
  arrange(time_stamp) 

t <- df_1$time_stamp
y <- df_1$x_UTM
f <- approxfun(t,y)

interp_df$x_UTM<-f(interp_df$time_stamp)
y <- df_1$y_UTM
f <- approxfun(t,y)
interp_df$y_UTM<-f(interp_df$time_stamp)
interp_df<-interp_df[c("x_UTM","y_UTM")]#only select x and y columns

```


#Assess numbers of clusters

```{r k-means clusters, include=FALSE,eval = FALSE}

silhouette_score <- function(k){
  km <- kmeans(interp_df, centers = k, nstart=10, iter.max = 300,algorithm = c("Lloyd"))
  ss <- silhouette(km$cluster, dist(interp_df))
  mean(ss[, 3])
}

k <- 2:10
avg_sil <- sapply(k, silhouette_score)
#plot(k, type='b', avg_sil, xlab='Number of clusters', ylab='Average Silhouette Scores', frame=FALSE)
optimal_k<- which.max(avg_sil)+1

km.out <- kmeans (interp_df, optimal_k, nstart = 10, iter.max = 300,algorithm = c("Lloyd"))

interp_df$cluster<-km.out$cluster

#select only clusters with at least two observations
interp_df<-interp_df %>% 
     group_by(cluster) %>% 
     filter(n() >1)

interp_df_sf = st_as_sf(interp_df, coords = c("x_UTM", "y_UTM"), crs = crs_utm)
interp_df_sf<-st_transform(interp_df_sf, crs_original)#
coords<-as.data.frame(st_coordinates(interp_df_sf))#extract coordinates
interp_df$lon<-coords$X
interp_df$lat<-coords$Y

wardpal <- colorFactor(viridis(5), interp_df$cluster)

 leaflet() %>%
    addTiles(urlTemplate = "https://mts1.google.com/vt/lyrs=s&hl=en&src=app&x={x}&y={y}&z={z}&s=G", attribution = 'Google') %>%
   #addProviderTiles(providers$CartoDB.Positron)%>% 
     addCircleMarkers(lng = ~lon,
              lat = ~lat,
              radius=5,
              popup = ~ as.character(cluster),
              label = ~as.character(cluster),
              fillColor=~wardpal(cluster),
              fillOpacity = 1,
              stroke=FALSE,
              data = interp_df) %>%
    addScaleBar()%>%
   addLegend(position="bottomright",pal=wardpal,values=interp_df$cluster)

```
 
 
#Create ellipses

```{r ellipse 95, include=FALSE,eval = FALSE}
 
interp_df$cluster<-as.factor(interp_df$cluster)

ellipse<-dataEllipse(interp_df$lon, interp_df$lat, levels=0.95,groups=interp_df$cluster)
id=names(ellipse)
ellipses <- do.call(rbind.data.frame, ellipse)#150976

f<-data.frame()

for ( i in seq_along(id)){
 f<- rbind( f,as.data.frame(rep(id[i],nrow(ellipse[[i]]))))
}

ellipses<-cbind(ellipses,f)
colnames(ellipses)<-c("x","y","id")

#make polygons, extract points not in ellipse
library(sf)
library(sfheaders)

ellipses_polygons <- sfheaders::sf_polygon(
  obj = ellipses
  , x = "x"
  , y = "y"
  , polygon_id = "id"
)
sf::st_crs( ellipses_polygons) <- 4326

plot(ellipses_polygons)

df_sf = st_as_sf(interp_df, coords = c("lon", "lat"), crs = 4326)
#df_sf_utm<-st_transform(df_sf, crs_utm)

points_in_polygons <-lengths(st_intersects(df_sf, ellipses_polygons)) > 0#
interp_df<-interp_df[c(points_in_polygons),]#remove points outside polygons

#calculate center for each ellipse 

centroids<-interp_df%>%
  group_by(cluster)%>%
  summarise(lon_mean=mean(lon),lat_mean=mean(lat))
centroids_sf = st_as_sf(centroids, coords = c("lon_mean", "lat_mean"), crs = crs_original)
centroids_utm<-st_transform(centroids_sf, crs_utm)

centroids_utm_buff<-centroids_utm%>%
  st_buffer(dist=centroids_buff) #polygons with potential mooring or anchoring sites generated.
plot(centroids_utm_buff)

st_write(centroids_utm_buff, "potential_mooring_sites.shp")

```

#Defining trips

```{r define trips, include=FALSE,eval = FALSE}

df_sf = st_as_sf(df, coords = c("x_UTM", "y_UTM"), crs = crs_utm)
points_in_polygons <-lengths(st_intersects(df_sf, centroids_utm_buff)) > 0#
df$in_polygon<-points_in_polygons

df$trip_id<-rleid(df$in_polygon)

#select points outside polygons
df<-df%>%
  filter(in_polygon==FALSE)
#proceed with code in Sup. mat XX - Step 7 - Remove suprious trips


```
